---
title: "Layers of plots"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Layers in `ggplot2`

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

---

# Outline

- grammar of graphics, again

- layer specifications


---

# Grammar of Graphics

A graphical representation (plot) consists of:

1. **mappings** (`aes`):  data variables are mapped to graphical elements

2. **layers**:  geometric elements (`geoms`, such as points, lines, rectangles, text, ...) and statistical transformations (`stats`, are identity, counts, bins, ...)

3. **scales**: map values in the data space to values in an aesthetic space (e.g. color, size, shape, but also position)

4. **coordinate system** (`coord`): normally Cartesian, but pie charts use e.g. polar coordinates

5. **facetting**: for small multiples (subsets) and their arrangement

6. **theme**: fine-tune display items, such as font and its size, color of background, margins, ...



---

# Layers

each layer has several parts, the two most important are:

- mapping: using the `aes()` function, we specify mappings between variables and aesthetics of the chart

- `data`: the dataset used in the layer 

`ggplot` is the layer that sets the defaults, <br>

`geom_XXX` creates a layer: for each aspect it either uses the default or has to specify its own

---

# Previous example

```{r, fig.width=9.5, fig.height = 3}
data(nasa, package="GGally")
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = date, y= temperature)) + geom_point()
```

---

# Connecting the points

```{r, fig.width=9.5, fig.height = 3}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = date, y= temperature)) + geom_point() +
  geom_line()
```

---

# Adding a trend line

```{r, fig.width=9.5, fig.height = 3}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = date, y= temperature)) + geom_point() +
  geom_line() +
  geom_smooth(method="lm", se=FALSE)
```



---

# Adding additional text

```{r, fig.width=9.5, fig.height = 3}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = date, y= temperature)) + geom_point() +
  geom_line() +
  geom_smooth(method="lm", se=FALSE) +
  geom_text(aes(label = time, x = date+weeks(4)))
```

---

# Labelling specific text

```{r, fig.width=9.5, fig.height = 3}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = date, y= temperature)) + geom_point() +
  geom_line() +
  geom_smooth(method="lm", se=FALSE) +
  geom_text(aes(label = time, x = date+weeks(4)),
            nasa %>% filter(id=="1-1", temperature > 298))
```

---

# Default versus Layer Specification

```{r, fig.width=9.5, fig.height = 3, message = FALSE}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = time, y= temperature)) + geom_point() +
  geom_line() +
  geom_text(aes(label=id, colour=id),  
            data=nasa %>% filter(id=="1-1", time==50))
```

---


```{r, fig.width=9.5, fig.height = 3, message = FALSE}
nasa %>% filter(id=="1-1") %>%
  ggplot(aes(x = time, y= temperature, colour=id)) + geom_point() +
  geom_line() +
  geom_text(aes(label=id), 
            data=nasa %>% filter(id=="1-1", time==50))
```


