---
title: "DS 202 - lab #4: Scraping (into) the Hall of Fame"
author: Heike Hofmann
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
# Outline

- example of scraping with selectors

---
# Data Background

The Lahman data package is based on [Sean Lahman](https://www.seanlahman.com/)'s  Baseball  [Database](https://www.seanlahman.com/baseball-archive/statistics/). 

The `HallOfFame` table is a part of this set of tables and  has [not been updated since 2018](https://github.com/chadwickbureau/baseballdatabank/commits/a0b17f910506f7bfc2be2f834c50d50beb884c8f/core/HallOfFame.csv?browsing_rename_history=true&new_path=contrib/HallOfFame.csv&original_branch=master)

## Baseball Reference 

The site baseball-reference.com has grown out of Sean Lahman's work and is now maintained independently. 

 Incidentally, it also has tables with Hall of Fame information, e.g. for 2022: 

https://www.baseball-reference.com/awards/hof_2022.shtml

---
# Scrape the data

Use the `rvest` package to download and parse data tables for Hall of Fame voting records. 

```{r message=FALSE}
url <- "https://www.baseball-reference.com/awards/hof_2022.shtml"

library(rvest)
site <- read_html(url)
```

The command `html_element` allows us to select based on css selectors (www3 school CSS)[https://www.w3schools.com/CSSref/css_selectors.php] or (CSS Diner)[https://flukeout.github.io/]

Load the baseball reference website in Chrome. Then use View > Developer > Inspect Elements.

What id should we use?

---

# BBWAA Table

table has id `hof_BBWAA`:

```{r}
site %>% html_element(css="#hof_BBWAA") %>% html_table() %>% head(3)
```

Solve the problem with the first row by writing the table into a temporary file and reading it back in.

---

# Write the table, read the table

```{r}
bbwaa <- site %>% html_element(css="#hof_BBWAA") %>% html_table() 
readr::write_csv(bbwaa, "temp.csv")
bbwaa <- readr::read_csv("temp.csv", skip = 1, show_col_types = FALSE)
head(bbwaa)
```


---

# Clean the data

What steps are necessary to get the scraped data into the shape as the `HallOfFame` data table:

```{r}
library(Lahman)
head(HallOfFame, 3)
```
How do we get the number of ballots and needed votes? - Go back to Chrome!

---

# Data Munging

```{r}
site %>% html_element(css="#hof_BBWAA_sh div") %>% html_text()
```
```{r}
bbwaa <- bbwaa %>% select(1:7) %>% 
  mutate(
  yearID = 2022,
  votedBy = "BBWAA",
  ballots = 394,
  needed = 296,
  votes = Votes,
  inducted = ifelse(votes >= 296, "Y", "N"),
  category = "Player",
  needed_note = NA
)
```

How do we get `playerID`? 

---

# Match `People` and `bbwaa` by players' names

```{r}
People <- People %>% mutate(
  Name = paste(nameFirst, nameLast)
)

bbwaa <- bbwaa %>% mutate(
  Name = gsub("X-", "", Name)
)
```

anti join: what will go wrong?

```{r}
bbwaa %>% anti_join(People %>% select(Name, playerID), by="Name")
```

---

# what's with Pierzynski?

```{r}
People %>% filter(nameLast=="Pierzynski") %>% select(Name, playerID)
```
There is an extra space!

Idea: replace all occurrences of . by a space ' ', then delete multiple spaces

---

# Space 

```{r}
People <- People %>% mutate(
  Name = gsub(".", " ", Name, fixed=TRUE),
  Name = gsub("  ", " ", Name)
)
bbwaa <- bbwaa %>% mutate(
  Name = gsub(".", " ", Name, fixed=TRUE),
  Name = gsub("  ", " ", Name)
)

bbwaa %>% anti_join(People %>% select(Name, playerID), by="Name")
```
Now we can join the data sets

---

# Getting the player ID

```{r}
bbwaa <- bbwaa %>% left_join(People %>% select(Name, playerID), by="Name")
```

# The right variables in the right order

```{r}
bbwaa %>% select(names(HallOfFame))
```




---
# Deliverable: data

```{r fig.width= 12, fig.height = 4, warning = FALSE}
HallOfFame <- rbind(HallOfFame, bbwaa %>% select(names(HallOfFame)))

HallOfFame %>% 
  ggplot(aes(x = yearID, fill = inducted)) +
  geom_bar() +
  xlim(c(1936, 2023))
```

---

# Still missing

- Veterans' votes for 2022

- Hall of fame data for 2021

- Hall of fame data for 2020

- Hall of fame data for 2019

Lots of repetition! 