---
title: "Functions in R"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
class: middle, inverse, center
# Function writing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
```

---

# Functions in R

- Have been using functions a lot, now we want to write them ourselves!

- Idea: avoid repetitive coding (errors will creep in)

- Instead: extract common core, wrap it in a function, make it reusable


---

# Structure of functions

- Name

- Input arguments

    - names, 

    - default values

- Body

- Output values

---

# A first function

```{r}
mymean <- function(x) {
	return(sum(x)/length(x))
}
```

```{r}
mymean(1:15)
mymean(c(1:15, NA))
```

---

# A first function (2)

```{r}
mymean <- function(x, na.rm=F) {
	if (na.rm) x <- na.omit(x)
	
	return(sum(x)/length(x))
}

mymean(1:15)
mymean(c(1:15, NA), na.rm=T)
```


---
class: inverse
# Your Turn: a scraper 

The package `rvest` allows us to download data from the baseball reference website `url` using the following lines of code:

```
library(rvest)
site <- read_html(url)
bbwaa <- site %>% html_element("#hof_BBWAA") %>% html_table()

head(bbwaa)
```

Write a function that uses the url as input argument, scrapes the data and returns it

Try out your function on the site
https://www.baseball-reference.com/awards/hof_2021.shtml

---

# Your turn - solution

```{r warning = FALSE, message = FALSE}
library(rvest)

bbwaa_scraper <- function(url) {
  site <- read_html(url)
  bbwaa <- site %>% html_element("#hof_BBWAA") %>% html_table()
  
  bbwaa
}

bbwaa_scraper("https://www.baseball-reference.com/awards/hof_2021.shtml")
```

---
class: inverse
# Your Turn: expanding 

Expand your function by a parameter `element` that enables you to download different pieces from a website.

Set `element` to "#hof_Veterans" and try out your function for the year 2022.


---
class: inverse
# Your Turn: a helper function

Write a helper function `dots_to_spaces` that takes as input a vector of characters (text), and returns as output the same vector in which all occurrences of '.' are replaced and all double spaces are reduced to one.

```{r, eval = FALSE}
dots_to_spaces <- function(x) {
  # body of the function
  
  # return cleaned up vector x
}
```

---
class: inverse, center, middle
# Always scrape data responsibly!
