---
title: "Graphics with ggplot2"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.width=3.5, fig.height=3.5)
```

## Data Exploration

- asking good questions is important for any data exploration: "your data won't speak unless you ask the right questions"

- a data exploration is a cycle of asking questions, looking at the answers, and deriving new questions from them

- for a sound exploration: start by checking individual variables, then move on to higher dimensions

---
class: center, middle
# An example exploration


---
class: inverse
# Your turn 


- Write down questions that you could answer with this data 4 minutes by yourself, then pair up for another 3 minutes, and we'll write ideas on the board

```{r}
library(classdata)
head(fbiwide)
```

---

# Plan for answers

- Explore how one (or more) variables are distributed: *barchart or histogram*
- Explore how two variables are related:  *scatterplot, boxplot, tile plot*
- Explore how two variables are related, conditioned on other variables: *facetting, color & other aesthetics*


We will be using the `ggplot2` framework for plotting

```{r}
library(ggplot2)
```

---

# Why `ggplot2`

- Wildly popular package for statistical graphics: over 2.5 million downloads from CRAN  (several thousand times per day)

- Developed by Hadley Wickham (An ISU Alumni)

- Designed to adhere to good graphical practices

- Constructs plots using the concept of layers

- Supports a wide variety plot types and extensions

- Ported to different languages, e.g. `ggpy`  (`seaborn` is also grammar of graphis based)
\medskip

- http://ggplot2.org/book/ or Hadley's book *ggplot2: Elegant Graphics for Data Analysis* for reference 


---

# `ggplot` Function

The `ggplot` function is the basic workhorse of ggplot2

- Produces all plot types available with ggplot2

- Allows for plotting options within the function statement

- Creates an object that can be saved

- Plot layers can be added to modify plot complexity

---

# `ggplot` Structure

The `ggplot` function has the basic syntax:

`ggplot(data, mappings) + geom_type(options)`

- data: dataset to be used

- mappings: determines which variables are connected to which plot elements, mappings are done with `aes()`

- type: determines type of the plot, e.g. `point`, `line`, `bar`

- options: there are so, so many options!

---

# Scatterplots in `ggplot2`

`aes` allows us to specify mappings; scatterplots need a mapping for `x` and a mapping for `y`:

```{r}
ggplot(data = fbiwide, aes(x = Burglary, y = Murder)) +
  geom_point()
```  

---
class: inverse
# Your turn 


- Draw a scatterplot of the number of burglaries by murders.

- Adjust the numbers of the above scatterplot to show log transformed numbers. How can we interpret the numbers on the axes?

- Draw a scatterplot of the log transformed number of burglaries by motor vehicle thefts.

---

# Revision - Interpreting Scatterplots

- Big patterns

    - Form and direction

    - Strength

- Small patterns

- Deviations from the pattern

    - Outliers
    
---

# Aesthetics

Can map other variables to size, colour, shape, ....

```{r}
ggplot(aes(x = log(Burglary), y = log(Motor.vehicle.theft),
           colour=Year), data=fbiwide) + geom_point()
```           

---

# Your turn 

- Draw a scatterplot of the log transformed number of burglaries by motor vehicle thefts. Map the state variable to colour. Why is this a terrible idea?

- Draw a scatterplot of the log transformed number of burglaries by motor vehicle thefts. Map Population to size. How do we interpret the output?

- Which other aesthetics are there? Have a look at the [RStudio cheat sheet on visualization](https://www.rstudio.com/resources/cheatsheets/) 


**Stretch goal**: Draw a histogram of the state populations. 

