---
title: "Reshaping data with tidyr - working with separate and unite"
author: "Heike Hofmann"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
# Outline

- cleaning data (first run)

- functions `separate` and `unite`


---
# parse_number 

the tidyverse package `readr` contains function `parse_number`

```{r}
library(readr)
x <- c("3.14", "5.2%", "$10", "5,321.00")
x
parse_number(x)
```
---
class: inverse
# Your Turn

Work on the previously created data frame `deaths` (Avengers).

```{r echo=FALSE, include=FALSE, message = FALSE}
library(tidyverse)
av <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/avengers/avengers.csv", stringsAsFactors = FALSE)
deaths <- av %>% 
  pivot_longer(starts_with("Death"), 
               names_to="Time", values_to = "Died") %>%
  select(URL, Name.Alias, Time, Died) %>%
  filter(Died %in% c("YES", "NO"))
```

- Use `parse_number` to extract from the variable `Time` a number. Inspect the result.

- Group by Avenger (use URL) and `Died`. Find the maximum of `Time`. Call the result `maxdeaths`. Interpret the resulting data set.

```{r, echo=FALSE}
deaths %>% 
  mutate(Time = parse_number(Time)) %>%
  group_by(URL, Died) %>% 
  summarize(Time = max(Time)) %>% 
  head(10)
```
---
class: inverse
# Your Turn

Work with the `maxdeaths` dataset. Find the frequency break down by Time and Died (`group_by`, `tally`).

```{r echo=FALSE, message=FALSE}
deaths %>% 
  mutate(Time = parse_number(Time)) %>%
  group_by(URL, Died) %>% 
  summarize(Time = max(Time)) %>% 
  group_by(Died, Time) %>%
  tally()
```
Use the same steps for all of the Avengers' returns and you have the basic info for all statements for lab #3.

---
# separate

Messy (2): Multiple variables are stored in one column

```{r, message=FALSE, warning = FALSE}
library(tidyverse)
df <- data.frame(x = c(NA, "a.b", "a.d", "b.c"))
df
df %>% separate(x, into = c("A", "B"))
```

---
class: inverse
# Your Turn (5 min)


The Iowa Data Portal is a wealth of information on and about the State of Iowa. 

The website 
[Liquor Sales](https://data.iowa.gov/Sales-Distribution/Iowa-Liquor-Sales/m3tr-qhgy) provides data on every liquor sale in a licensed store in Iowa. The code below reads (part of) the data into an R session.

```
url <- "https://github.com/ds202-at-ISU/materials/blob/master/03_tidyverse/data/ames-liquor.rds?raw=TRUE"
download.file(url, "ames-liquor.rds", mode="wb")
ames <- readRDS("ames-liquor.rds")
```


---

# Problems with the data

- `Date` is text, in the format of Month/Day/Year (Messy 2)

- Store location is a textual expression of form `POINT (`...`)` and geographic latitude and longitude. (Messy 2)


no Messy 1? - problems of type Messy 1 are typically hard to detect and often up to interpretation/dependent on the analysis to be done.

---
class: inverse
# Your Turn (10 min)


- Check the help for the function `parse_number` in the `readr` package and use it on the store location. What result do you get?

- Use `separate` to split the variable for store location into longitude and latitude.
 (Hint - you might need several steps of `separate`)

- Make sure that latitude and longitude are numeric variables. 

---
class: inverse
# Your Turn - exploration (10 mins)

Use `dplyr` functionality to answer the following questions:

  1. What is the total amount spent on Liquor Sales?

  2. What is the single largest sale (in volume/in dollar amount)?

  4. Plot geographic longitude and latitude. Where are liquor sales in Ames happening? 


